---
title: "Assignment 4"
author: "Ann Eitrheim"
output: html_notebook
---

#Importing the Data
```{r, results='hide', warning = FALSE}
files <- list.files(path = "~/Documents/Time_Series_Analysis/Assignment_4/Data",
                    pattern = "*.xls", full.names = T)
files <- append(files[2:16],files[1])
library(readxl)
df <- sapply(files, read_excel, simplify=FALSE)
library(dplyr)
df <- bind_rows(df, .id = "id")
df <- select(df, "X__2","X__4")
colnames(df) <- c("hour","count")
library(tidyverse)
df <- df %>% filter(str_detect(hour, "00"))
datess <- substr(files, 79,85)
datecol = c()
for (i in 1:16) {datecol = append(datecol,rep.int(datess[i],24))}
df <- add_column(df, "dates" = datecol, .after = 0)
df <- as.data.frame(df)
df$count <- as.numeric(df$count)
```

```{r}
head(df)
rm(datecol,datess,files,i) #cleaning up the global environment
```

#The daily data is from Illinois Dept of Transporation (IDOT)
for I80E 1EXIT (the 2nd data column) note each data point is an hourly count of the number of vehicles at a specific location on I80E. Use the daily data for last 2 weeks of June 2013 to develop an ARIMA forecasting model. Objective is to forecast the hourly counts for July 1. The actual data file for July 1 is included for you to test your estimate.

```{r}
train <- df[1:360,3]
julyAct <- df[361:384,3]
```

#Part 1 - Use ARIMA(p,d,q) model to forecast. Find the model returned by R auto.arima(). Change the values of p and q and determine the best model using AICc and BIC. Do AICc and BIC select the same model as the best model?

```{r}
hourly <- ts(train, start = c(16,1), end = c(30,24),frequency = 24)
ts.plot(hourly)
```

```{r}
#Best model using AICc
require(forecast)
hourlyAICc <- auto.arima(train,
                         stepwise=F, #changing values of p and q in a grid pattern
                                     #to determine the best model
                         seasonal = F, #we are asked for ARIMA so I restricted the
                                       #search to non-seasonal models
                         ic = 'aicc', #telling the function to select based on the
                                      #best AICc 
                         trace = T) #to see what models are considered and the approx
                                    #AICc each model gives
summary(hourlyAICc)
```

Output for the best model based on AICc is ARIMA(2,0,3) with non-zero mean  
AIC=4455.56   AICc=4455.88   BIC=4482.77

```{r}
#Best model using BIC
hourlyBIC <- auto.arima(hourly,
                        stepwise=F,
                        seasonal = F,
                        ic = 'bic',  #telling to select based on the best BIC
                        trace = T)
summary(hourlyBIC)
```

Output for the best model based on BIC is ARIMA(2,0,2) with non-zero mean  
AIC=4456.52   AICc=4456.76   BIC=4479.83

We get different models if we select based on the best AICc or BIC, as they select different values for q. Selecting based on the best AICc results in q equaling 3, yet q is 2 if we selected based off of best BIC.

```{r}
#Forecasting the next 24 hours (i.e. July 1) for both models
julyforecast.hourlyAICc <- forecast(hourlyAICc, h=24)
julyforecast.hourlyBIC <- forecast(hourlyBIC, h=24)

x <- data.frame("Hour" = seq(1,24,1),
                "Actuals" = julyAct,
                "AICc" = round(as.numeric(julyforecast.hourlyAICc$mean,1)),
                "BIC" = round(as.numeric(julyforecast.hourlyBIC$mean,1)),
                "AICc minus BIC" = round(as.numeric(julyforecast.hourlyAICc$mean) -
                                           as.numeric(julyforecast.hourlyBIC$mean),1))

print.data.frame(x)

c("AICc.SSE"=sum((julyAct-julyforecast.hourlyAICc$mean)^2),
  "BIC.SSE"=sum((julyAct-julyforecast.hourlyBIC$mean)^2))
```

The models have predictions that are close to one another, but the model selected based on BIC has a slighly lower sum of squared errors than AICc for July 1. Yet the model selected on the best AICc has the lowest RMSE on the training set (114.97 vs 115.44).

# Part 2 - Use day of the week seasonal ARIMA(p,d,q)(P,Q,D)s model to forecast for July 1 (which is a Monday) - note use the hourly data

```{r}
weekly <- ts(train, end = c(2,180),frequency = 24*7)
ts.plot(weekly)
weeklySARIMA <- auto.arima(weekly,
                           stepwise=F,
                           max.order = 20) #to set max value of p+q+P+Q
summary(weeklySARIMA)
```

Output for the best model is ARIMA(0,1,2)(0,1,0)[168]   
AIC=2249.31   AICc=2249.44   BIC=2259.07

```{r}
#Forecasting the next 24 hours (i.e. July 1)
julyforecast.weeklySARIMA <- forecast(weeklySARIMA, h=24)
plot(julyforecast.weeklySARIMA, xlim=c(2.2, 3.2)) #showing one week
```

# Part 3 - Use hour of the day seasonal ARIMA (p,d,q)(P,D,Q)s model to forecast for the hours 8:00, 9:00, 17:00 and 18:00 on July 1

```{r}
#creating a seasonal ARIMA model for the hour of day
hourlySARIMA <- auto.arima(hourly)
summary(hourlySARIMA)
```

Output for the best model is ARIMA(2,0,2)(2,1,0)[24] 

```{r}
julyforecast.hourlySARIMA <- forecast(hourlySARIMA, h=24)
data.frame("Actuals" = julyAct[c(8,9,17,18)],
           "Hourly.Model" = julyforecast.hourlySARIMA$mean[c(8,9,17,18)],
            row.names = c('8:00', '9:00', '17:00', '18:00'))
```

# Part 4 - For the July 1 8:00, 9:00, 17:00 and 18:00 forecasts, which model is better (part 2 or part 3) ?

```{r}
#Comparing the acutal values with the predicted model
data.frame("Actuals" = julyAct[c(8,9,17,18)],
           "Weekly.Model" = julyforecast.weeklySARIMA$mean[c(8,9,17,18)],
           "Hourly.Model" = julyforecast.hourlySARIMA$mean[c(8,9,17,18)],
           row.names = c('8:00', '9:00', '17:00', '18:00'))
```

```{r}
#Looking the prediction error (how far away the prediction is from the actual value)
Weekly.Error <- julyAct[c(8,9,17,18)] - julyforecast.weeklySARIMA$mean[c(8,9,17,18)]
Hourly.Error <- julyAct[c(8,9,17,18)] - julyforecast.hourlySARIMA$mean[c(8,9,17,18)]
data.frame("Weekly.Error" = round(abs(Weekly.Error),2),
           "Hourly.Error" = round(abs(Hourly.Error),2),
           row.names = c('8:00', '9:00', '17:00', '18:00'))
```

```{r}
#Finding the sum of squred errors for the predictions
#(Only looking at July 1st 8:00, 9:00, 17:00 and 18:00 hours)
c('Weekly SSE'=round(sum(Weekly.Error^2),1),
  'Hourly SSE'=round(sum(Hourly.Error^2),1))
```

The weekly SARIMA model is better as it has a much lower sum of squared errors; seasonality based on day of week is better at forecasting the hours 8:00, 9:00, 17:00, and 18:00 than seasonality based on hour of day. 

Additionally, the weekly SARIMA has a lower RSME on the training set (60.97 vs. 101.95). While the hourly SARIMA model did a better job predicting traffic at 17:00, the other predictions are over 10x further off than the weekly model. It looks like the weekly model (blue line) does a better job of following the movement of the actual values (black line) in the graph below, for the exception of hours 19:00 and 20:00. This makes sense because a week prior (June 24) there was a 0 count at 19:00 and 253 at 20:00. There might have been construction or some other exogenous factor.

```{r, echo = FALSE}
plot(seq(1,24), julyAct, col = 'black', 
     main = "Forecasted Values versus Actual for July 1", 
     type = 'o', xlab = 'Hour', ylab = "Count", axes = FALSE)
axis(side=1, at=seq(1,24,2))
axis(side=2, at=seq(0, 1250, 300))
box()
lines(seq(1,24), julyforecast.weeklySARIMA$mean, col="blue")
lines(seq(1,24), julyforecast.hourlySARIMA$mean, col="orange")
legend(8.5,500, legend=c("Actual Values", "Weekly SARIMA", "Hourly SARIMA"), 
       col=c("black", "blue", "orange"), lty=1:3, cex=0.8)
```
